---
title: "Analysis NN: "
format:
  html:
    embed-resources: true
---
```{r}
library(data.table)
library(dplyr)
library(knitr)
```
# Overview
Calculate quantitative inhibition metrics from GWA data
```{r}
pheno_path <- "data/processed/phenotypes/pheno.df.rda"

# toxicant condition metadata
tox_meta_fn <- "data/processed/tox_data/tox_metadata.csv"
```
# Inputs
Load the phenotype data
```{r}
# load phenotype data
load(pheno_path)

dplyr::glimpse(pheno.df)
```

Load the toxicant metadata (for nice_drug_label2)
```{r}
tox_meta <- data.table::fread(tox_meta_fn)
dplyr::glimpse(tox_meta)
```

Extract the relevant columns:
* `median_wormlength_um`: Median wormlenght for a well
* `drug`: Toxicant
* `strain`: Strain name
* `control_pheno`: The control phenotype for that strain
* `well.id`: The well identifier 

```{r}
clean_pheno_df <- pheno.df %>%
  dplyr::select(
    drug,
    strain,
    median_wormlength_um,
    control_pheno,
    well.id
  ) 
```

# Validate the data
These data contain well-level estimates of toxicant responses. For each unique `strain`, we expected to 26 or fewer unique `drug` values, and 12 wells per `drug`/`strain` combination

```{r}
strain_drug_counts <- clean_pheno_df %>%
  dplyr::group_by(strain) %>%
  dplyr::summarise(
    n_drugs = dplyr::n_distinct(drug),
    .groups = "drop"
  )
knitr::kable(head(strain_drug_counts))
```


Check the number of unique drugs per strain and the number of unique control well phenotypes per strain/drug combination
```{r}
validation_summary <- clean_pheno_df %>%
  dplyr::group_by(strain, drug, .drop = FALSE) %>%
  dplyr::summarise(
    n_wells = dplyr::n_distinct(well.id),
    n_distinct_control_pheno = dplyr::n_distinct(control_pheno),
    .groups = "drop"
  )

knitr::kable(head(validation_summary))
```

# Calculate inhibition metrics

## Step 1: Calculate % growth inhibition for each well (rowwise)

Calculate the percent growth inhibition for each well using the paired `control_pheno` value. The calculation is performed rowwise to preserve the pairing between each toxicant measurement and its corresponding control phenotype.

Formula: `% growth inhibition = 100 × (1 – (median_wormlength_um / control_pheno))`

```{r}
well_inhibition <- clean_pheno_df %>%
  dplyr::mutate(
    pct_growth_inhibition = 100 * (1 - (median_wormlength_um / control_pheno))
  )

dplyr::glimpse(well_inhibition)
```

## Step 2: Summarize % growth inhibition per strain/toxicant

For each strain in each toxicant condition, calculate the mean % growth inhibition across all wells. Also report the number of distinct `control_pheno` values and wells per strain/drug combination.

```{r}
strain_inhibition <- well_inhibition %>%
  dplyr::group_by(drug, strain) %>%
  dplyr::summarise(
    pct_growth_inhibition = mean(pct_growth_inhibition, na.rm = TRUE),
    mean_length_exposed = mean(median_wormlength_um, na.rm = TRUE),
    mean_control_pheno = mean(control_pheno, na.rm = TRUE),
    n_control_pheno_values = dplyr::n_distinct(control_pheno),
    n_wells = dplyr::n(),
    .groups = "drop"
  )
```

```{r}
knitr::kable(head(strain_inhibition))
```

Report the distribution of distinct control phenotype values per strain/drug:
```{r}
knitr::kable(as.data.frame(table(strain_inhibition$n_control_pheno_values)),
             col.names = c("n_control_pheno_values", "Frequency"))
```

Preview the distribution of inhibition values across all strain/toxicant combinations:
```{r}
ggplot2::ggplot(strain_inhibition, ggplot2::aes(x = pct_growth_inhibition)) +
  ggplot2::geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "% Growth Inhibition Distribution Across Strain/Toxicant Combinations",
    x = "% Growth Inhibition",
    y = "Count"
  )
```

```{r}
summary(strain_inhibition$pct_growth_inhibition)
```

### Investigate any negative inhibition values

```{r}
negative_inhibition <- strain_inhibition %>%
  dplyr::filter(pct_growth_inhibition < 0)
```

```{r}
knitr::kable(negative_inhibition)
```

## Step 3: Summarize % growth inhibition per toxicant condition (mean ± SD across strains)

For each toxicant, calculate the mean and standard deviation of % growth inhibition across all strains used in the GWAS.

```{r}
toxicant_inhibition_summary <- strain_inhibition %>%
  dplyr::group_by(drug) %>%
  dplyr::summarise(
    mean_pct_inhibition = mean(pct_growth_inhibition, na.rm = TRUE),
    sd_pct_inhibition = sd(pct_growth_inhibition, na.rm = TRUE),
    n_strains = dplyr::n(),
    .groups = "drop"
  )

dplyr::glimpse(toxicant_inhibition_summary)
```

Display the summary table:
```{r}
knitr::kable(toxicant_inhibition_summary %>% dplyr::arrange(desc(mean_pct_inhibition)))
```

## Step 4: Join with toxicant metadata for nice_drug_label2

Join the inhibition summary with the toxicant metadata to include the `nice_drug_label2` values for integration with Figure 2.

```{r}
# Select drug and nice_drug_label2 from toxicant metadata
drug_labels <- tox_meta %>%
  dplyr::select(drug, nice_drug_label2) %>%
  dplyr::distinct()

# Join with inhibition summary
toxicant_inhibition_final <- toxicant_inhibition_summary %>%
  dplyr::left_join(drug_labels, by = "drug") %>%
  # Clean up nice_drug_label2 to replace uM with µM for plotting
  dplyr::mutate(nice_drug_label2 = stringr::str_replace(nice_drug_label2, pattern = "uM", replacement = "µM"))

dplyr::glimpse(toxicant_inhibition_final)
```

```{r}
knitr::kable(toxicant_inhibition_final %>% dplyr::arrange(desc(mean_pct_inhibition)))
```

# Save output

Save the toxicant inhibition summary for use in Figure 2.

```{r}
output_path <- "data/processed/phenotypes/toxicant_inhibition_summary.csv"
data.table::fwrite(toxicant_inhibition_final, output_path)
message("Saved toxicant inhibition summary to: ", output_path)
```

