---
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(pheatmap)
library(cowplot)
library(glue)
library(here)
```

```{r}
# Source the portable project setup function
script_dir <- tryCatch(
  dirname(normalizePath(knitr::current_input())),
  error = function(e) getwd()
)

# Try multiple possible locations for the setup file
setup_file_paths <- c(
  file.path(script_dir, "..", "bin", "setup_project_here.R"), # Standard location relative to script
  file.path(getwd(), "bin", "setup_project_here.R"), # If we're in project root
  file.path(getwd(), "projects", "ToxinGWAS_Manuscript", "bin", "setup_project_here.R") # If in projects structure
)

setup_file <- NULL
for (path in setup_file_paths) {
  if (file.exists(path)) {
    setup_file <- path
    break
  }
}

if (!is.null(setup_file)) {
  source(setup_file)
  project_root <- setup_toxin_project_here()
} else {
  stop(
    "Could not find setup_project_here.R in any of the expected locations: ",
    paste(setup_file_paths, collapse = ", ")
  )
}

if (is.null(project_root)) {
  stop("Could not locate ToxinGWAS_Manuscript project root.")
}

```

```{r}
source("bin/outs.R")
source("bin/materials_key.R")
```

```{r}
#| include: false
# Load correlation data
pheno_corr_all <- readRDS(file.path(project_root, "data/processed/phenotypes/pheno_corr_all.Rds"))
tox_data <- data.table::fread(file.path(project_root, "data/processed/tox_data/con_metadata.csv"))
strain_means_pivot_matrix <- readRDS(file.path(project_root, "data/processed/phenotypes/strain_means_matrix.rds"))
```

Report out the average correlation across all toxicant conditions
```{r}
init_corr_coef <- pheno_corr_all$r

corr_coef_df <- as_tibble(init_corr_coef, rownames = "trait1") %>%
  pivot_longer(-trait1, names_to = "trait2", values_to = "r_value")

# calculate the average correlation
pheno_corr_all_avg <- corr_coef_df %>%
  dplyr::filter(trait1 != trait2) %>%
  dplyr::summarise(
    mean_r = mean(r_value, na.rm = TRUE),
    sd_r = sd(r_value, na.rm = TRUE),
    q1_r = quantile(r_value, probs = 0.25, na.rm = TRUE),
    q3_r = quantile(r_value, probs = 0.75, na.rm = TRUE),
    upper_threshold_r = q3_r + 1.5 * (q3_r - q1_r),
    lower_threshold_r = q1_r - 1.5 * (q3_r - q1_r)
  )
pheno_corr_all_avg
````

To annotate the cells of the clustered heatmap reprocess the correlation data after performing the clustering so that the annotation matricies align with the clusterd conditions

```{r}
# perform clustering on pairwise Spearman correlation coefficients
clust <- pheatmap::pheatmap(init_corr_coef)

# get the clustered order of the toxicants
desired_order <- rownames(init_corr_coef)[clust$tree_row$order]

# reshape the strain matrix to match cluster order
strain_means_pivot_matrix2 <- strain_means_pivot_matrix[, desired_order]

pheno_corr_all2 <- psych::corr.test(
  strain_means_pivot_matrix2,
  use = "pairwise",
  adjust = "bonferroni",
  method = "spearman"
)

# get the reordered correlation coefficients and p-values
pheno_corr_all_r <- pheno_corr_all2$r
pheno_corr_all_p <- pheno_corr_all2$p
pheno_corr_all_p_symbol <- ifelse(round(pheno_corr_all2$p, digits = 3) <= 0.05, "*", " ")
```

Check to identify significant correlations:

Lower triangle with un-corrected p-valyes
```{r}
# pull the lower and top diagonals for detailed analysis
# 299 pos, 26 neg
# lower raw
# 185 pos, 2 neg sig raw
lower_r_raw <- pheno_corr_all2$r
lower_r_raw[upper.tri(lower_r_raw, diag = T)] <- NA_real_
lower_r_proc <- as_tibble(lower_r_raw) %>%
  dplyr::mutate(y_trait = row.names(lower_r_raw)) %>%
  tidyr::pivot_longer(cols = -y_trait, names_to = "x_trait", values_to = "r") %>%
  dplyr::filter(!is.na(r)) %>%
  dplyr::mutate(
    neg_n = sum(r < 0),
    pos_n = sum(r > 0)
  )

lower_p_raw <- pheno_corr_all2$p
lower_p_raw[upper.tri(lower_p_raw, diag = T)] <- NA_real_
lower_p_proc <- as_tibble(lower_p_raw) %>%
  dplyr::mutate(y_trait = row.names(lower_p_raw)) %>%
  tidyr::pivot_longer(cols = -y_trait, names_to = "x_trait", values_to = "raw_p") %>%
  dplyr::filter(!is.na(raw_p)) %>%
  dplyr::mutate(`raw_p<0.05` = ifelse(raw_p < 0.05, T, F))

lower_proc <- dplyr::left_join(lower_r_proc, lower_p_proc) %>%
  dplyr::mutate(
    pos_sig_n = sum(r > 0 & `raw_p<0.05` == T),
    neg_sig_n = sum(r < 0 & `raw_p<0.05` == T)
  )

```

Upper triangle with Bonferroni corrected p-values
```{r}
# top or bonferroni corrected
# 88 pos, 0 neg sig
top_r_raw <- pheno_corr_all2$r
top_r_raw[lower.tri(top_r_raw, diag = T)] <- NA_real_
top_r_proc <- as_tibble(top_r_raw) %>%
  dplyr::mutate(y_trait = row.names(top_r_raw)) %>%
  tidyr::pivot_longer(cols = -y_trait, names_to = "x_trait", values_to = "r") %>%
  dplyr::filter(!is.na(r)) %>%
  dplyr::mutate(
    neg_n = sum(r < 0),
    pos_n = sum(r > 0)
  )

top_p_raw <- pheno_corr_all2$p
top_p_raw[lower.tri(top_p_raw, diag = T)] <- NA_real_
top_p_proc <- as_tibble(top_p_raw) %>%
  dplyr::mutate(y_trait = row.names(top_p_raw)) %>%
  tidyr::pivot_longer(cols = -y_trait, names_to = "x_trait", values_to = "BF_p") %>%
  dplyr::filter(!is.na(BF_p)) %>%
  dplyr::mutate(`BF_p<0.05` = ifelse(BF_p < 0.05, T, F))

top_proc <- dplyr::left_join(top_r_proc, top_p_proc) %>%
  dplyr::mutate(
    pos_sig_n = sum(r > 0 & `BF_p<0.05` == T),
    neg_sig_n = sum(r < 0 & `BF_p<0.05` == T)
  )
top_proc
```

```{r}
#| include: false
# Create the color palette and annotation data


# Set color palette for annotations
class.pal.df <- tibble::tibble(
  col = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#D55E00"),
  big_class = c("Flame Retardant", "Fungicide", "Herbicide", "Insecticide", "Metal")
)

# filter the meta data to include only toxicant and join to the class palette
tox_cols <- tox_data %>%
  # remove the controls
  dplyr::filter(big_class != "Control") %>%
  dplyr::select(drug, nice_drug_label2, big_class, moa_class) %>%
  # Clean up to have special character for plotting
  dplyr::mutate(nice_drug_label2 = stringr::str_replace(nice_drug_label2, pattern = "uM", replacement = "µM")) %>%
  # join to the class palette
  dplyr::left_join(., class.pal.df, by = c("big_class" = "big_class")) %>%
  dplyr::select(drug, nice_drug_label2, big_class, moa_class, col) %>%
  dplyr::rename(MoA = moa_class, Use = big_class) %>%
  # In moa_class column replace Other|unknown with "Other / unknown"
  dplyr::mutate(MoA = stringr::str_replace(MoA, pattern = "Other\\|unknown", replacement = "Other / unknown"))

# Create a mapping from drug to nice_drug_label2
drug_to_label <- setNames(tox_cols$nice_drug_label2, tox_cols$drug)


# Update row and column names of correlation matrices
rownames(pheno_corr_all_r) <- drug_to_label[rownames(pheno_corr_all_r)]
colnames(pheno_corr_all_r) <- drug_to_label[colnames(pheno_corr_all_r)]
rownames(pheno_corr_all_p_symbol) <- drug_to_label[rownames(pheno_corr_all_p_symbol)]
colnames(pheno_corr_all_p_symbol) <- drug_to_label[colnames(pheno_corr_all_p_symbol)]

moa_anno_df <- tox_cols %>%
  # set the row names to the nice drug label
  tibble::column_to_rownames(var = "nice_drug_label2") %>%
  # rename the big class column to Class
  # dplyr::rename(Class = Use) %>%
  # remove the drug and color columns
  dplyr::select(-drug, -col)
# In moa_class column replace Other|unknown with "Other / unknown"
# dplyr::mutate(moa_class = stringr::str_replace(moa_class, pattern = "Other|unknown", replacement = "Other / unknown")) %>%
# dplyr::rename(MoA = moa_class, Use = Class)


# Create annotation colors
cols <- unique(class.pal.df$col)
names(cols) <- unique(class.pal.df$big_class)
anno_cols <- list(Use = cols)
```

```{r}
#| include: false
create_phenotype_heatmap <- function(similarity_matrix, con_metadata_df, anno_cols, significance_matrix) {
  # Perform hierarchical clustering on rows to get the dendrogram
  row_dist <- dist(similarity_matrix)
  row_hclust <- hclust(row_dist)

  # Use the same clustering order for both rows and columns
  # (since it's a symmetric similarity matrix)
  p <- pheatmap::pheatmap(
    similarity_matrix,
    annotation_row = con_metadata_df,
    annotation_colors = anno_cols,
    width = 7.5,
    height = 6,
    color = colorRampPalette(c("navy", "white", "darkred"))(1000),
    breaks = seq(-1, 1, length.out = (1000 + 1)),
    show_colnames = TRUE,
    fontsize_number = 12,
    display_numbers = significance_matrix,
    cluster_rows = row_hclust,
    cluster_cols = FALSE,
    cutree_rows = 2,
    silent = TRUE
  )
  return(p)
}
```


```{r}
#| label: fig-phenotype-correlations
#| fig-width: 7.5
#| fig-height: 6
phenotype_correlations <- create_phenotype_heatmap(
  similarity_matrix = pheno_corr_all_r,
  con_metadata_df = moa_anno_df,
  significance_matrix = pheno_corr_all_p_symbol,
  anno_cols = anno_cols
)
phenotype_correlations
```

Save publication figure
```{r}
save_plot(
  tplot = phenotype_correlations,
  fn_list = figure_fns$phenotype_correlations,
  w_in = 7.5,
  h_in = 5
)
```